---
title: "Reproducible Research - Peer Assessment2"
author: "robertwcw"
date: "9/21/2020"
output: 
  html_document: 
    keep_md: yes
    toc_depth: 4
    fig_caption: yes
---
## SYNOPSIS

  Disaster triggered by storm, tornado and a host of other weather phenomenons may lead to public health issues and dire economic consequences. The extent, severity, and impact of injuries/fatalities and economic losses are largely determined by the amount of energy unleashed by the types of events in the weather related disasters. This data analysis attempts to address the concerns **WHERE** the events took place and **WHAT** types of events gave rise to deleterious social impact aftermath with respect to economic and public health. Base on the US National Oceanic and Atmospheric Administration (NOAA) storm database, this project makes use of relevant metrics of the number of human casualty, financial losses in properties/crops damages in relation to events types across the US from year 1950 to year 2011.  

  Off-shore affiliated territories of the USA are protectorates not under direct administration of the USA, therefore shall be excluded from this analysis work.  
&nbsp;

#### Initialize R workspace
```{r setup, tidy=TRUE}
knitr::opts_chunk$set(echo = TRUE)
Sys.setenv(TZ = "UTC")      # set global TZ to UTC for POSIXt class

.Rfliburl <- "https://raw.githubusercontent.com/robertwcw/Rflib/master"
source(file.path(.Rfliburl,"getRflib.R"),local = TRUE)
source(getRflib("is.defined.R"),local = TRUE)
source(getRflib("myplclust.R"),local = TRUE)
source(getRflib("strCap.R"),local = TRUE)

if (!requireNamespace("data.table",quietly = TRUE)) install.packages("data.table")
library(data.table)
 
if (!requireNamespace("dplyr",quietly = TRUE)) install.packages("dplyr")
library(dplyr, warn.conflicts = FALSE)

if (!requireNamespace("lubridate",quietly = TRUE)) install.packages("lubridate")
library(lubridate, quietly = TRUE, warn.conflicts = FALSE)

if (!requireNamespace("maps",quietly = TRUE)) install.packages("maps")
library(maps)

if (!requireNamespace("lattice", quietly  = TRUE)) install.packages("lattice")
library(lattice)

if (!requireNamespace("scales", quietly  = TRUE)) install.packages("scales")
library(scales)

if (!requireNamespace("stringr", quietly  = TRUE)) install.packages("stringr")
library(stringr)

# 'udf.num2gcs' user-defined function to convert LATITUDE & LONGITUDE coordinate values in integer type to nnn.dddd geographic coordinate system format 
udf.num2gcs <- function(X, y) { # y is number of significant digits.
        if (is.na(X) | is.null(X) | !is.numeric(X)) return(X)
        x <- as.character(X)
        l <- nchar(x)
        ifelse(l < 3, y <- l,
               ifelse(l == 3, y <- 1,
                      ifelse(l == 4, y <- 2,
                             ifelse(l > 4, y <- y, y <- 0))))
        y <- l - y
        X <- as.numeric(paste0(x,"e",-y))
        X
}

# 'udf.stripsign' user-defined function to strip -ve sign from numeric value
udf.stripsign <- function(x) {
        if (!is.na(x) & !is.null(x) & is.numeric(x)) {
                x <- x * sign(x)
        }
        x
}

# 'udf.prependsign' user-defined function to prepend -ve sign to numeric value
udf.prependsign <- function(x) {
        if (!is.na(x) & !is.null(x) & is.numeric(x)) {
                x <- x * sign(-x)
        }
        x
}

# 'udf.str2time' user-defined function to put TIME string into proper format
udf.str2time <- function(x) {
        l <- nchar(x)
        if (l %in% c(3,4)) {
                z <- substr(strptime(sprintf("%04s", x), "%H%M"), 12, 19)
                ifelse(is.na(z) | z == "",
                       x <- "00:00:00",
                       x <- substr(strptime(sprintf("%04s", x), "%H%M"), 12, 19))
        } else
        if (l == 11) {
                z <- substr(strptime(x, "%I:%M:%S %p"), 12, 19)
                ifelse(is.na(z),
                       x <- substr(strptime(x, "%H:%M:%S"), 12, 19),
                       ifelse(z == "",
                              x <- substr(strptime(x, "%H:%M:%S %p"), 12, 19),
                              x <- z))
        }
        x
}

# 'udf.expont' user-defined function for converting number scale character to proper scientific notation.
udf.expont <- function(x, y) {
        x <- ifelse(x %in% c("1","2","3","4","5","6","7","8","9"), paste0("e+0",x),
                    ifelse(x == "H", "e+02",
                           ifelse(x == "K", "e+03",
                                  ifelse(x == "M", "e+06",
                                         ifelse(x == "B", "e+09", "e+0")))))
        as.numeric(paste0(as.character(y),x))
}

# list of states of the USA (exclude off-shore affiliated territories).
mystates <- c(unique(state.fips$abb), "AK", "HI") 

# set up colour palette.
set.seed(2)
i <- sample.int(502, 256)
i <- sample(i, 66)
mypal <- colors(TRUE)[i]

# snapshot of default graphic parameters.
par.def <- par(no.readonly = TRUE) 

```
&nbsp;

#### Data Loading

  Loading into R workspace the compressed *raw* comma-delimited text file in bz2 format downloaded from the course website, as data.table class object with naming of the data object derived from the source file programmatically.

1) obtain the URL for NOAA Stormdata data set from the course website.
2) set up the mechanism to download the raw data file onto local storage.
3) uncompressed the downloaded raw file and extract a comma-delimited text file.
4) read the comma-delimited text file into R workspace as data table for efficiency.  

```{r data.load, tidy=TRUE}
# loading data downloaded from course website into R workspace 
#   user  system elapsed 
# 53.457   1.190  57.097 

# set data source url to course website
fileurl <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
datadir <- paste(".", "data", sep = "/")
filetmp <- tempfile()
filenam <- character()

download.file(fileurl, filetmp)

if (!dir.exists(datadir)) {
        dir.create(datadir)
}

# extract source data file name as name of work data object in workspace 
filels <- strsplit(basename(fileurl), "%2F", fixed = TRUE)[[1]]
for (i in 1:length(filels)) 
        {
        filein <- strsplit(filels[i], ".", fixed = TRUE)[[1]]
        if (length(filein) > 1) 
                {
                filenam[i] <- filein[1]
                filesrc <- bzfile(filetmp, open = "r")
                assign(filenam[i], data.table(read.csv(filesrc)))
                flush(filesrc)
                }
        }
close(filesrc)

filenam <- filenam[which(!is.na(filenam))]

unlink(filetmp)
rm(fileurl,filetmp,filesrc,filels,filein)
```
&nbsp;

#### Exploratory Data Analysis

```{r eda.0, tidy=TRUE}
for (i in 1:length(filenam)) {
        print(paste(i, ":",filenam[i], "(",floor(object.size(get(filenam[i]))/10e+05),"MB)"))
}
```

  **`r length(filenam[1])`** data `r ifelse(length(filenam) > 1, "sets were", "set was")` extracted and loaded into R workspace from NOAA sourced data downloaded from the course website. (number in bracket depicts memory footprint).  

Structure of **`r filenam[1]`** `r class(get(filenam[1]))[1]`.

```{r eda.1, tidy=TRUE}
str(get(filenam[1]))
```

  Visual examination of the **`r filenam[1]`** `r class(get(filenam[1]))[1]` indicating there are a number of data vectors having incorrect atomic vector class, entail further processing and transformation to change their atomic class to the appropriate ones, which is compulsory to enable correct downstream analytic processing. As understood, the raw data set is quite a bit, *dirty*.  
&nbsp;


## Data Processing

  For the purpose of this analysis work, which is to answer the pertinent questions about the weather events of *WHAT and WHERE*; we will proceed to spawn a new data.table object called **stormsub** from existing raw data set `r filenam[1]` with subset of 11 variables, plus a few more working-variables derived from existing raw data set. The following is the subset of variables grabbed from the existing raw data set **`r filenam[1]`**:  
```{r data.proc, tidy=TRUE}
v <- c(5,7,8,23,24,25,26,27,28,32,33)
va <- array(dim = c(length(v),2))
for (i in 1:length(v)) {
        va[i,1] <- names(get(filenam[1]))[v[i]]
        va[i,2] <- class(names(get(filenam[1]))[v[i]])
}
print(va)
```
column 1: Name of Variable  
column 2: Data Type  
&nbsp;


**EV_TIMESTAMP**  *working-variable*  
  Generate a time-stamp in POSIXct format, computed base on existing raw data BGN_DATE & BGN_TIME for each event as it was recorded in the raw data set. 
```{r data.proc.1, tidy=TRUE}
d1 <- sub(" .*", "", StormData$BGN_DATE)
d2 <- gsub("[oO]", "0", StormData$BGN_TIME) # rectify typo error of 'O' for '0'
d2 <- sub("12:00:00 AM", "00:00:00", d2) # rectify non-POSIXt compliant time fmt
d2 <- sub("00:00:00 AM", "00:00:00", d2) # rectify non-POSIXt compliant time fmt
d2 <- unlist(lapply(d2, udf.str2time)) # unify various time format to "hh:mm:ss"

StormData <- StormData %>% mutate(EV_TIMESTAMP = as.POSIXct(strptime(paste(d1,d2), "%m/%d/%Y %H:%M:%S")))

rm(d1, d2)
```
&nbsp;

**LONGITUDE** & **LATITUDE**  
  The geographic coordinates (x,y) used in mapping weather events onto the map of USA to visually tell user where were the events took place. **LATITUDE** and **LONGITUDE** data in their current form are not immediately use-able by any mapping system, must go through a conversion process to transform them into geographic coordinate system configuration to be readily recognizeable by base R mapping tools or any other mapping systems.
```{r data.proc.2, tidy=TRUE}
# Reformat LATITUDE vector to geographic coordinate system format nnn.ddd
numsign <- sign(StormData$LATITUDE)
lat <- sapply(StormData$LATITUDE, udf.stripsign)
lat <- sapply(lat, udf.num2gcs, 2) # '2' is number of significant digits
StormData$LATITUDE <- lat * (numsign * numsign) # LATs belong to northern hemisphere 

# Reformat LONGITUDE vector to geographic coordinate system format nnn.ddd
numsign <- sign(StormData$LONGITUDE)
lon <- sapply(StormData$LONGITUDE, udf.stripsign)
lon <- sapply(lon, udf.num2gcs, 3) # '3' is number of significant digits
StormData$LONGITUDE <- lon * (numsign * -numsign) # LONs belong to western hemisphere

rm(lon, lat, numsign)  
```
&nbsp;

**EVTYPE**  
  Data values captured in EVTYPE in its current raw format is quite a bit "dirty" and mostly present itself in human natural language construct than conforming to standard category coding format useful for information processing using computers. To enable simpler data processing logic coding to avoid unnecessary error during analytic processing run, we trim un-needed white-spaces from within **EVTYPE** vector then convert to upper case for conformity with the requirement of downstream analytic processing. Please refer to **EVCLASS** for further explanation.
```{r data.proc.3, tidy=TRUE}
# Trim whitespace & convert to upper case.
StormData$EVTYPE <- toupper(str_squish(StormData$EVTYPE))
```
&nbsp;

**STATE**  
**COUNTY**  
**PROPDMGEXP**  
**CROPDMGEXP**  
  Trim un-needed embedded white-space, if any, then convert these data values to upper case words, again for conformity.
```{r data.proc.4, tidy=TRUE}
StormData$STATE <- toupper(str_squish(StormData$STATE))
StormData$COUNTY <- toupper(str_squish(StormData$COUNTY))
StormData$PROPDMGEXP <- toupper(str_squish(StormData$PROPDMGEXP))
StormData$CROPDMGEXP <- toupper(str_squish(StormData$CROPDMGEXP))
```
&nbsp;

**EVCLASS**  *working-variable*  
  Re-classification of event type, to standard category coding format to overcome the non-standardize text format present in **EVTYPE**, to ensure they are in conformity with the requirement of downstream analytic processing.
```{r data.proc.5, tidy=TRUE}
# Add NEW column EVCLASS to StormData
StormData <- StormData %>% mutate(EVCLASS = character(1))

# Re-classification of weather events type.
# https://www.weather.gov/iwx/events
# A - TORNADOs of all kinds 
# B - EXTREME TEMPERATURES on both end of the scale
# C - SNOWY & ICY conditions
# D - HIGH WINDS & STORM
# E - RAIN & FLOOD
# F - HAILing condition
# G - CLOUD condition that may lead to storm or tornado event
# H - FIRE 
# I - LIGHTNING
# X - EXCLUDED list - weather events to be excluded from processing
# Z - OTHER events not classified 

# Array Index 24, 25 & 26 of the ev.class array are the special designated classes.
# DO NOT ALTER the hard-designated values in index 24, 25 & 26, except the EXCLUSION keyword for weather events at index [24,3].
# Array Index 24 - EXCLUSION class, use for assigning weather events that are to be excluded from analytic processing. Change the search keyword of the weather event in the character string at index[24,3] as required.
# Array Index 25 - reserved.
# Array Index 26 - OTHER class, use for classifying weather events not classified by the regular class codes from index 1 to 23. Do NOT alter.
# DO NOT assign empty string value to the array cells, otherwise will result in unexpected outcome.
# DO NOT ALTER the sequence order of the default class codes and search keywords (array index 1 ~ 23).
ev.class <- array(dim = c(26,3))
ev.class[1,1] <- "A"        # CLASS CODE
ev.class[1,2] <- "TORNADO"  # CLASS LABEL
ev.class[1,3] <- "TORNADO"  # SEARCH KEYWORD
ev.class[2,1] <- "B"
ev.class[2,2] <- "EXTREME.TEMP"
ev.class[2,3] <- "HEAT|HOT|DRY|DROUGHT|COLD|CHILL|FROST|TEMP"
ev.class[3,1] <- "C"
ev.class[3,2] <- "SNOW.ICE"
ev.class[3,3] <- "SNOW|ICE|WINTER|BLIZZARD|FREEZ(E|ING)|SLEET|ICY"
ev.class[4,1] <- "D"
ev.class[4,2] <- "WIND.STORM"
ev.class[4,3] <- "WIND|STORM|GUSTNA|HURRICAN|TYPHOON|TURBULENCE|BURST|DUST|TSTM"
ev.class[5,1] <- "E"
ev.class[5,2] <- "RAIN.FLOOD"
ev.class[5,3] <- "FLOOD|WET|RAIN|SHOWER|HEAVY|PRECIPITAT"
ev.class[6,1] <- "F"
ev.class[6,2] <- "HAIL"
ev.class[6,3] <- "HAIL"
# ev.class[7,1] <- "G"  # commented - shall go into OTHERS class automatically
# ev.class[7,2] <- "CLOUD"
# ev.class[7,3] <- "CLOUD|FUNNEL|WALL"
ev.class[8,1] <- "H"
ev.class[8,2] <- "WILD.FIRE"
ev.class[8,3] <- "FIRE"
ev.class[11,1] <- "I"
ev.class[11,2] <- "LIGHTNING"
ev.class[11,3] <- "LIGHTNING"
ev.class[24,1] <- "X"         # EXCLUSION class. Hard-designated. DO NOT ALTER.
ev.class[24,2] <- "EXCLUDED"  # class label. Hard-designated. DO NOT ALTER.
ev.class[24,3] <- "SUMMARY"   # EXCLUSION keyword, change as necessary.
ev.class[26,1] <- "Z"         # Hard-designated. DO NOT ALTER.
ev.class[26,2] <- "OTHERS"    # class label. Hard-designated. DO NOT ALTER.
ev.class[26,3] <- "OTHERS"    # OTHER class. DO NOT ALTER.

P <- which(!is.na(ev.class[1:23,1]))
N <- length(P)

# EXCLUDED - weather events to exclude
evclass.rowid <- grep(ev.class[24,3], StormData$EVTYPE)
evclass.evtyp <- StormData[evclass.rowid,]$EVTYPE
evclass <- cbind(evclass.rowid, evclass.evtyp)
StormData <- StormData %>% 
        mutate(EVCLASS = replace(EVCLASS, evclass.rowid, ev.class[24,1]))

# weather events defined by the regular search keywords
for (i in 1:N) {
        evclass.rowid <- grep(ev.class[P[i],3], StormData$EVTYPE)
        evclass.rowid <- evclass.rowid[!(evclass.rowid %in% as.integer(evclass[,1]))]
        evclass.evtyp <- StormData[evclass.rowid,]$EVTYPE
        evclass.tmp <- cbind(evclass.rowid, evclass.evtyp)
        evclass <- rbind(evclass, evclass.tmp)
        StormData <- StormData %>% 
              mutate(EVCLASS = replace(EVCLASS, evclass.rowid, ev.class[P[i],1]))
}

# OTHERS - weather events not classified by the regular class codes
evclass.rowid <- as.integer(row.names(StormData))
evclass.rowid <- evclass.rowid[!(evclass.rowid %in% as.integer(evclass[,1]))]
evclass.evtyp <- StormData[evclass.rowid,]$EVTYPE
evclass.tmp <- cbind(evclass.rowid, evclass.evtyp)
evclass <- rbind(evclass, evclass.tmp)
StormData <- StormData %>% 
        mutate(EVCLASS = replace(EVCLASS, evclass.rowid, ev.class[26,1]))

rm(evclass.tmp, evclass.evtyp, evclass.rowid, evclass)
```
&nbsp;

**PROPDMGS**  *working-variable*  
  Amount of property damages derived from **PROPDMG** augmented by number scale in **PROPDMGEXP**.

**CROPDMGS**  *working-variable*  
  Amount of crop damages derived from **CROPDMG** augmented by number scale in **PROPDMGEXP**.

**stormsub**  *working-data set*  
  Comprising subset of the weather event data spawned from the raw data set **StormData**, instituting the basis of all subsequent analytic processing and reporting. 
```{r data.proc.6, tidy=TRUE}
# subset of StormData across the USA (exclude off-shore territories).
stormsub <- StormData %>% 
        select(EV_TIMESTAMP, EVCLASS, EVTYPE, LONGITUDE, LATITUDE, STATE, COUNTY,
               FATALITIES, INJURIES, PROPDMG, PROPDMGEXP, CROPDMG, CROPDMGEXP,
               REFNUM) %>%
        filter(STATE %in% mystates & !(EVCLASS %in% c("X","Z"))) %>% 
        arrange(EV_TIMESTAMP) %>%
        mutate(STATE = factor(STATE),
               LONGITUDE = round(LONGITUDE, 1),
               LATITUDE = round(LATITUDE, 1),
               EVCLASS = factor(EVCLASS, labels = ev.class[P,2]),
               PROPDMGS = mapply(udf.expont, PROPDMGEXP, PROPDMG),
               CROPDMGS = mapply(udf.expont, CROPDMGEXP, CROPDMG)) 
```
  The affiliated territories of the USA are excluded from **stormsub** data set.
&nbsp;

## Results

<!-- weather events frequency aggregate overview barchart -->
```{r data.analytic, fig.cap="Figure 1: Overview of Weather Events across USA", message=FALSE, tidy=TRUE, fig.width=13, fig.height=12} 
par(par.def)

pal <- mypal[seq(length(mypal), by = -2, length.out = N)]

# set up plotting layout and graphics parameters 
par(las = 1, cex.axis = 0.80, cex.lab = 0.85, ann = FALSE)
nf <- layout(matrix(c(1,2,
                      1,2,
                      3,3,
                      3,3,
                      3,3,
                      4,4,
                      4,4,
                      4,4), nrow = 8, byrow = TRUE))

# weather events aggregate frequency by type across the USA.
evcnt1 <- stormsub %>%
        select(STATE,EVCLASS,EVTYPE,EV_TIMESTAMP) %>%
        group_by(EVCLASS) %>%
        summarise(EVCNT = n())

# plotting barchart for weather events aggregate by type across the USA.

par(mai = c(0.25, 0.92, 0.25, 0.10))

barplot(evcnt1$EVCNT, horiz = TRUE, border = NA, col = pal, names.arg = unique(evcnt1$EVCLASS))

mtext("Overall Event Aggregate Frequency Across USA (1950 ~ 2011)", font = 2, side = 3, line = 0.25, adj = 1, cex = 0.75)
mtext("(US Affiliated Territories Not Included)", side = 3, line = -0.75, adj = 1, cex = 0.65)

maxtik <- max(axTicks(1))
m <- barplot(evcnt1$EVCNT, horiz = TRUE, plot = FALSE)
for (i in 1:length(m)) {
        j <- ifelse(evcnt1$EVCNT[i] >= maxtik, -2.25, 0.25)
        text(evcnt1$EVCNT[i], m[i], labels = formatC(evcnt1$EVCNT[i], big.mark = ","), pos = 4,  offset = j, cex = 0.70)
}

# barchart plot for weather events manifested in casualty and property/crop loss by type across the USA.
evcnt2 <- stormsub %>%
        select(EV_TIMESTAMP, STATE, EVCLASS, EVTYPE, 
               FATALITIES, INJURIES, PROPDMGS, CROPDMGS) %>%
        filter(FATALITIES > 0 | INJURIES > 0 | PROPDMGS > 0 | CROPDMGS > 0) %>%
        group_by(EVCLASS) %>%
        summarise(EVCNT = n())

# plotting barchart for weather events frequency aggregate manifested in casualty and property/crop loss by type across the USA.

par(mai = c(0.25,0.10,0.25,0.82), yaxt = "n")

barplot(evcnt2$EVCNT, horiz = TRUE, border = NA, col = pal, names.arg = unique(evcnt2$EVCLASS))

mtext("Events Manifested in Casualty & Economic Loss", font = 2, side = 3, line = 0.25, adj = 1, cex = 0.75)
mtext("Frequency aggregated across USA (1950 ~ 2011)", font = 2, side = 3, line = -0.75, adj = 1, cex = 0.75)
mtext("(US Affiliated Territories Not Included)", side = 3, line = -1.75, adj = 1, cex = 0.65)

maxtik <- max(axTicks(1))
m <- barplot(evcnt2$EVCNT, horiz = TRUE, plot = FALSE)
for (i in 1:length(m)) {
        j <- ifelse(evcnt2$EVCNT[i] >= maxtik, -2.25, 0.25)
        text(evcnt2$EVCNT[i], m[i], labels = formatC(evcnt2$EVCNT[i], big.mark = ","), pos = 4,  offset = j, cex = 0.70)
}

# hazardous weather events in terms of casualty count and economic loss
# casualty & property/crop loss breakdown by event type of individual-state
evcnt3 <- stormsub %>%
        select(STATE, EVCLASS, EVTYPE, FATALITIES,INJURIES, PROPDMGS,CROPDMGS) %>%
        group_by(STATE, EVCLASS) %>%
        summarise(ECONLOSS = (sum(PROPDMGS) + sum(CROPDMGS)) / 1e+09,
                  CASUALTI = (sum(FATALITIES) + sum(INJURIES)) / 1e+03)

par(mai = c(0.25, 0.92, 0.25, 0.82), cex.axis = 0.75, ann = FALSE, yaxt = "s")

barplot(CASUALTI ~ EVCLASS + STATE, data = evcnt3,
        border = NA,
        col = pal)

title(ylab = "Casualty Count ( '000 )", cex.lab = 0.90)
mtext("Casualty by Event Type of Individual-State (1950 ~ 2011)", font = 2, side = 3, line = -3.00, adj = 0, padj = 0, cex = 0.75)

barplot(ECONLOSS ~ EVCLASS + STATE, data = evcnt3,
        border = NA,
        col = pal)

title(ylab = "Economic Loss ( $Bn )", cex.lab = 0.90)
mtext("Economic Loss by Event Type of Individual-State (1950 ~ 2011)", font = 2, side = 3, line = -3.00, adj = 1, padj = 0, cex = 0.75)

```


  Figure 1 above consisting 4 barcharts, each chart illustrating different dimensions but correlated weather events data in the USA. Related events are grouped together for this reporting purpose, e.g. all wind related events such as high wind, storm, thunderstorm, gustnado ect are grouped under **WIND.STORM** type. **RAIN.FLOOD** type consists of all water related events such as heavy rains and floods or events that lead to flooding.

Qeustion 1.  
Which types of events are most harmful with respect to population health?

**Tornado**, **Storm**, **Flood** and **Hail** (appear in the order of severity in terms of number of casualty) are most hazardous with respect to population health.

Question 2.
Which types of events have the greatest economic consequences?

**Storm**, **Flood** and **Tornado** (appear in the order of severity of damages caused measured in US$) are most devastated in terms of economic consequences.

  Conclusion: Not in any particular order, **Tornado**, **Storm** and **Flood** are most hazardous weather with respect to casualty count and economic loss in the history of USA.  
&nbsp;

<!-- trends of weather events across the usa from 1950 ~ 2011 -->
```{r data.analytic2, fig.cap="Figure 2: Trends of Weather Events", message=FALSE, tidy=TRUE, fig.width=12, fig.height=8}
par(par.def)

evcnt4 <- stormsub %>% 
        select(EV_TIMESTAMP, EVCLASS, 
               FATALITIES, INJURIES, PROPDMGS, CROPDMGS) %>%
        arrange(EV_TIMESTAMP) %>%
        group_by(YEARS = year(EV_TIMESTAMP), EVCLASS) %>%
        summarise(EVCNT = n(),
                  CASUALTI.mean = mean(FATALITIES + INJURIES),
                  ECONLOSS.mean = mean(PROPDMGS + CROPDMGS),
                  CASUALTI.sum = sum(FATALITIES + INJURIES),
                  ECONLOSS.sum = sum(PROPDMGS + CROPDMGS))

xyp <- xyplot(EVCNT ~ YEARS | EVCLASS, data = evcnt4,
              groups = EVCLASS,
              layout = c(2,4),
              pch = 19,
              col = pal,
              strip = FALSE,
              strip.left = TRUE,
              par.strip.text = list(cex = 0.6),
              scales = list(cex = 0.6, y = list(log = "e")),
              xlab = list(label = "Years of Observation", cex = 0.8),
              ylab = list(label = "Frequency of Weather Event (log)", cex = 0.8))

xyp <- update(xyp, panel = function(x, y, ...) {
        panel.xyplot(x, y, ...)
        panel.lmline(x, y, col = rgb(0,0,1, alpha = 0.5), lwd = 2)
        panel.abline(h = mean(y), col = rgb(1,0,0, alpha = 0.5), lwd = 2, lty = 2)
        })

print(xyp)
```


  From Figure 2 above, the frequency of occurrence of each event type are generally on the up trend since 1950 till 2011, except Lightning which was trending downward ever slightly. Closer inspection of the scatter plot indicates incident data of some event types e.g. **Extreme.Temp**, **Snow.Ice**, **Rain.Flood**, **Wild.Fire** and **Lightning** only began in 1993. 
  
  According to National Weather Service website <https://www.weather.gov/timeline>, in 1989 an eight year plan was announced for the modernization and restructuring of NWS agency. This massive overhaul changed the way the agency operates resulting in improved capabilities in data collection, hence multitude of data were available beginning 1993. However, different length of observation data does not invalidate earlier consensus that the rate of occurrence of all measured events are trending in the same direction and getting more rapid, except lightning.  
&nbsp;

<!-- maps of the usa with weather events mapping -->
```{r data.analytic3, fig.cap=paste("Figure 3: Locale of Most Hazardous Events (1990 ~ 2010)"), message=FALSE, tidy=TRUE, fig.width=12, fig.height=8}
# Mapping locations of Tornado, Storm & Flood events for year 1990 ~ 2010
par(par.def)

ev.klass <- c(1,4,5) # TORNADO, WIND.STORM & FLOOD

w <- which(year(stormsub$EV_TIMESTAMP) %in% c(1990:2010) & 
                   as.numeric(stormsub$EVCLASS) %in% ev.klass)

nf <- layout(matrix(c(1,1,1,1,2,2,
                      1,1,1,1,2,2,
                      1,1,1,1,2,2,
                      1,1,1,1,2,2,
                      1,1,1,1,3,3,
                      5,5,4,4,3,3,
                      5,5,4,4,3,3), nrow = 7, byrow = TRUE))

# Map of USA mainland
lon <- c(-130.00, -66.00)
lat <- c(25.00, 50.00)
# mymap0 <- map("world", regions = ".*usa", xlim = lon ,ylim = lat, interior = TRUE, plot = FALSE)
# plot(mymap0, type = "l", lwd = 0.5, cex.axis = 0.8, main = "USA mainland")
mymap0 <- map("state", interior = TRUE, plot = FALSE)
par(mai = c(0.25, 0.32, 0.25, 0.25))
plot(mymap0, type = "l", lwd = 0.5, ann = FALSE)
title(main = "USA mainland")
# coordinates of weather events for USA mainland
ev.w.gcs <- stormsub %>%
        select(STATE, COUNTY, EVCLASS, LONGITUDE, LATITUDE, 
               FATALITIES, INJURIES, PROPDMGS, CROPDMGS) %>%
        filter(as.numeric(rownames(stormsub)) %in% w &
                ((LONGITUDE > lon[1] & LONGITUDE < lon[2]) & 
                 (LATITUDE > lat[1] & LATITUDE < lat[2])) & 
               (FATALITIES+INJURIES > 0 | PROPDMGS+CROPDMGS > 0)) %>%
        arrange(LONGITUDE, LATITUDE) %>%
        group_by(EVCLASS, LONGITUDE, LATITUDE) %>%
        summarize(EVCNT = n(),
                  FATALITI = sum(FATALITIES),
                  INJURIES = sum(INJURIES),
                  PROPDMGS = sum(PROPDMGS),
                  CROPDMGS = sum(CROPDMGS))
points(x = ev.w.gcs$LONGITUDE, y = ev.w.gcs$LATITUDE, pch = 20, cex = 0.7, 
       col = alpha(pal[ev.klass], 0.2))
legend(x = -75, y = 25, ev.class[ev.klass,2], fill = pal[ev.klass], 
       border = pal[ev.klass], bty = "n", x.intersp = 1/2, y.intersp = 1, 
       xjust = 0, yjust = 0, text.font = 2, cex = 0.8)


# Map of Alaska
lon <- c(-180.00, -129.00)
lat <- c(50.33, 72.83)
mymap0 <- map("world", regions = ".*usa", xlim = lon ,ylim = lat, plot = FALSE)
par(mai = c(0.25, 0.32, 0.25, 0.25))
plot(mymap0, type = "l", lwd = 0.5, ann = FALSE)
title(main = "Alaska")
# coordinates of events for Alaska
ev.w.gcs <- stormsub %>%
        select(STATE, COUNTY, EVCLASS, LONGITUDE, LATITUDE, 
               FATALITIES, INJURIES, PROPDMGS, CROPDMGS) %>%
        filter(as.numeric(rownames(stormsub)) %in% w &
                ((LONGITUDE > lon[1] & LONGITUDE < lon[2]) & 
                 (LATITUDE > lat[1] & LATITUDE < lat[2])) &
                 (FATALITIES+INJURIES > 0 | PROPDMGS+CROPDMGS > 0)) %>%
        arrange(LONGITUDE, LATITUDE) %>%
        group_by(EVCLASS, LONGITUDE, LATITUDE) %>%
        summarize(EVCNT = n(),
                  FATALITI = sum(FATALITIES),
                  INJURIES = sum(INJURIES),
                  PROPDMGS = sum(PROPDMGS),
                  CROPDMGS = sum(CROPDMGS))
points(x = ev.w.gcs$LONGITUDE, y = ev.w.gcs$LATITUDE, pch = 20, cex = 1.1, 
       col = alpha(pal[ev.klass], 0.5))

# Map of Hawaii
lon <- c(-177.37, -153.48)
lat <- c(17.92, 29.45)
mymap0 <- map("world", regions = ".*usa", xlim = lon ,ylim = lat, plot = FALSE)
par(mai = c(0.27, 0.32, 0.25, 0.25))
plot(mymap0, type = "l", lwd = 0.5, ann = FALSE)
title(main = "Hawaii")
# coordinates of events 
ev.w.gcs <- stormsub %>%
        select(STATE, COUNTY, EVCLASS, LONGITUDE, LATITUDE, 
               FATALITIES, INJURIES, PROPDMGS, CROPDMGS) %>%
        filter(as.numeric(rownames(stormsub)) %in% w &
                ((LONGITUDE > lon[1] & LONGITUDE < lon[2]) & 
                 (LATITUDE > lat[1] & LATITUDE < lat[2])) &
                 (FATALITIES+INJURIES > 0 | PROPDMGS+CROPDMGS > 0)) %>%
        arrange(LONGITUDE, LATITUDE) %>%
        group_by(EVCLASS, LONGITUDE, LATITUDE) %>%
        summarize(EVCNT = n(),
                  FATALITI = sum(FATALITIES),
                  INJURIES = sum(INJURIES),
                  PROPDMGS = sum(PROPDMGS),
                  CROPDMGS = sum(CROPDMGS))
# ev.w.gcs <- subset(stormsub, as.numeric(rownames(stormsub)) %in% w &
#                            ((LONGITUDE > lon[1] & LONGITUDE < lon[2]) &
#                                     (LATITUDE > lat[1] & LATITUDE < lat[2])),
#                    c(LONGITUDE, LATITUDE)
#                    )
points(x = ev.w.gcs$LONGITUDE, y = ev.w.gcs$LATITUDE, pch = 20, cex = 1.1, 
       col = alpha(pal[ev.klass], 0.5))


par(mai = c(0.27, 0.62, 0.25, 0.25), ann = FALSE)

plot(log(CASUALTI.sum) ~ YEARS, 
     data = subset(evcnt4, as.numeric(EVCLASS) %in% ev.klass),
     pch = 19, col = pal[ev.klass], cex = 0.8)
title(main = "Number of Casualty", ylab = "(log scale)")

plot(log(ECONLOSS.sum) ~ YEARS, 
     data = subset(evcnt4, as.numeric(EVCLASS) %in% ev.klass),
     pch = 19, col = pal[ev.klass], cex = 0.8)
title(main = "Economic Loss", ylab = "(log scale)")
```


  Figure 3 above illustrates the locale of the 3-most hazardous event type in the USA for period 1990 ~ 2010. Interesting phenomenon observed here in the USA for the period where the 3-most harmful events mostly took place in the eastern half of the continent base on the geographic coordinate for each event as it occurred.

  For the same period, scatter plots for Economic Loss on the left and Number of Casualty on the right at the bottom of figure 3 yield an interesting situation where Economic Loss is rising while Number of Casualty is declining, as oppose to the general consensus of rising trend in general evident in figure 2 above.  
&nbsp;

```{r house.keeping, include=FALSE}
rm(lat, lon, nf, m, v, w, i, j, maxtik, va, xyp, ev.w.gcs, mymap0)
```


