---
title: "Reproducible Research - Peer Assessment2"
author: "robertwcw"
date: "9/21/2020"
output: 
  html_document: 
    fig_caption: yes
    keep_md: yes
---
## SYNOPSIS

Natural disaster triggered by storms and abnormal weather phenomenons may lead to public health issues and dire economic consequences. The extent, severity, and impact of injuries/fatalities and economic losses are largely determined by the amount of energy unleashed by the types of events in the weather related disasters. This data analysis attempts to address the concerns **WHERE** the events took place and **WHEN** the disasters struck by **WHAT** types of events give rise to deleterious social impact aftermath with respect to economic and public health. Base on the US National Oceanic and Atmospheric Administration (NOAA) storm database, this project makes use of relevant metrics of the number of human injuries/fatalities, financial losses in properties/crops damages in relation to events types across the US from year 1950 to year 2011.  
&nbsp;

#### Set up R workspace environment
```{r setup, message=TRUE, echo=FALSE, results="hide"}
knitr::opts_chunk$set(echo = TRUE)
Sys.setenv(TZ = "UTC")      # set global TZ to UTC for POSIXt class

.Rfliburl <- "https://raw.githubusercontent.com/robertwcw/Rflib/master"
source(file.path(.Rfliburl,"getRflib.R"),local = TRUE)
source(getRflib("is.defined.R"),local = TRUE)
source(getRflib("strCap.R"),local = TRUE)
# source(getRflib("myplclust.R"),local = TRUE)

library(data.table)
# library(lubridate)
library(dplyr)
library(maps)
# library(xts)

# 'udf.num2gcs' user-defined function to convert LATITUDE & LONGITUDE coordinate values in integer type to nnn.dddd geographic coordinate system format 
udf.num2gcs <- function(X, y) { # y is number of significant digits.
        if (!is.na(X) & !is.null(X) & is.numeric(X)) {
                x <- as.character(X)
                l <- nchar(x)
                ifelse(l < 3, return(as.numeric(x)),
                       ifelse(l == 3, y <- 1,
                              ifelse(l == 4, y <- 2,
                                     ifelse(l > 4, y <- y, y <- 0))))
                y <- l - y
                X <- as.numeric(paste0(x,"e",-y))
        }
        return(X)
}

# 'udf.stripsign' user-defined function to strip -ve sign from numeric value
udf.stripsign <- function(x) {
        if (!is.na(x) & !is.null(x) & is.numeric(x)) {
                x <- x * sign(x)
        }
        return(x)
}

# 'udf.prependsign' user-defined function to prepend -ve sign to numeric value
udf.prependsign <- function(x) {
        if (!is.na(x) & !is.null(x) & is.numeric(x)) {
                x <- x * sign(-x)
        }
        return(x)
}

# 'udf.str2time' user-defined function to put TIME string into proper format
udf.str2time <- function(x) {
        l <- nchar(x)
        x <- case_when(
                l %in% c(3,4) ~ {
                        z <- substr(strptime(sprintf("%04s", x), "%H%M"), 12, 19)
                        ifelse(is.na(z) | z == "", 
                               "00:00:00",
                               substr(strptime(sprintf("%04s", x), "%H%M"), 12, 19))
                },
                l == 11 ~ {
                        z <- substr(strptime(x, "%I:%M:%S %p"), 12, 19)
                        ifelse(is.na(z),
                               substr(strptime(x, "%H:%M:%S"), 12, 19),
                               ifelse(z == "",
                                      substr(strptime(x, "%H:%M:%S %p"), 12, 19),
                                      z))
                },
                TRUE ~ x
        )
        return(x) 
}
```
&nbsp;

#### Data Loading

Loading *raw* comma-delimited text file data set downloaded from course website into R workspace as data.table class object with naming of the data object derives from website data source programmatically, by-passing manual data staging as much as possible during loading phase.

```{r data.load, cache=TRUE, echo=TRUE, results="hide"}
# loading data downloaded from course website into R workspace 

# set data source url to course website
fileurl <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
datadir <- paste(".", "data", sep = "/")
filetmp <- tempfile()
filenam <- character()

download.file(fileurl, filetmp)

if (!dir.exists(datadir)) {
        dir.create(datadir)
}

# extract source data file name as name of work data object in workspace 
filels <- strsplit(basename(fileurl), "%2F", fixed = TRUE)[[1]]
for (i in 1:length(filels)) {
        filein <- strsplit(filels[i], ".", fixed = TRUE)[[1]]
        if (length(filein) > 1) {
                filenam[i] <- filein[1]
                filesrc <- bzfile(filetmp, open = "r")
                assign(filenam[i], data.table(read.csv(filesrc)))
                flush(filesrc)
        }
}
close(filesrc)

filenam <- filenam[which(!is.na(filenam))]

# # save an archive copy of StormData data.table to local `r datadir`
# for (i in 1:length(filenam)) {
#         fileout <- paste(datadir, filenam[i], sep = "/")
#         fileout <- paste(fileout, "csv", "bz2", sep = ".")
#         filesrc <- bzfile(fileout, open = "w")
#         write.csv(get(filenam[i]), filesrc, row.names = FALSE)
#         flush(filesrc)
# }
# close(filesrc)
# rm(fileout)
# 
# read StormData from saved archive copy data file on local disk
# filesrc <- paste(datadir, "StormData.csv.bz2", sep = "/")
# filesrc <- bzfile(filesrc, open = "r") # establish a connection to local file
# StormData <- data.table(read.csv(filesrc, skip = 0))
# flush(filesrc)
# close(filesrc)

unlink(filetmp)
rm(fileurl,filetmp,filesrc,filels,filein)
```
&nbsp;

#### Exploratory Data Analysis

```{r eda.0}
for (i in 1:length(filenam)) {
        print(paste(i, ":",filenam[i], "(",floor(object.size(get(filenam[i]))/10e+05),"MB)"))
}
rm(i)
```
**`r length(filenam)`** *raw* data `r ifelse(length(filenam) > 1, "sets were", "set was")` loaded into R workspace from NOAA sourced data downloaded from the course website. (number in bracket depicts memory footprint).

Structure of data object **`r filenam[1]`**.
```{r eda.1}
str(get(filenam[1]))
```
**`r length(names(get(filenam[1])))`** columns found in the *raw* data set **`r filenam[1]`**. 

<!-- For this data analysis work, we shall create a working data set called **stormdata** consisting 18 data columns grabbed from the *raw* data set **`r filenam[1]`**. (see attached list) -->

```{r data.proc, cache=TRUE, echo=TRUE}
# column number of the selected variable
v <- c(1,2,3,4,5,6,7,8,12,13,19,20,21,22,23,24,25,26,27,28,32,33,34,35,37)
names(get(filenam[1]))[v]

#    user  system elapsed 
# 601.279 116.964 720.355 
d1 <- sub(" .*", "", StormData$BGN_DATE)
d2 <- gsub("[oO]", "0", StormData$BGN_TIME) # rectify typo error of 'O' for '0'
d2 <- sub("12:00:00 AM", "00:00:00", d2) # rectify non-POSIXt compliant time fmt
d2 <- sub("00:00:00 AM", "00:00:00", d2) # rectify non-POSIXt compliant time fmt
d2 <- unlist(lapply(d2, udf.str2time)) # unify various time format to "hh:mm:ss"

StormData <- StormData %>% mutate(EV_TIMESTAMP = as.POSIXct(strptime(paste(d1,d2), "%m/%d/%Y %H:%M:%S")))

rm(d1,d2)

# # # NOT USED
# # prepend -ve sign to LONGITUDE values belong to the western hemisphere
#   StormData$LONGITUDE <- sapply(StormData$LONGITUDE, udf.prependsign)

# reformat LATITUDE vector to geographic coordinate system format nnn.ddd
  numsign <- sign(StormData$LATITUDE)
  lat <- sapply(StormData$LATITUDE, udf.stripsign)
  lat <- sapply(lat, udf.num2gcs, 2) # param '2' is number of significant digits
  StormData$LATITUDE <- lat * (numsign * numsign) # LATs belong to northern hemisphere 

#    user  system elapsed 
# 39.976   0.271  40.281 
# reformat LONGITUDE vector to geographic coordinate system format nnn.ddd
  numsign <- sign(StormData$LONGITUDE)
  lon <- sapply(StormData$LONGITUDE, udf.stripsign)
  lon <- sapply(lon, udf.num2gcs, 3) # param '3' is number of significant digits
  StormData$LONGITUDE <- lon * (numsign * -numsign) # LONs belong to western hemisphere

rm(lon, lat, numsign)  

# trim whitespaces on both ends of EVTYPE and transform to uppercase
StormData$EVTYPE <- toupper(trimws(StormData$EVTYPE))
# trim redundant intermediate whitespaces of length(n) to length(1)
repeat {
        StormData$EVTYPE <- sub("  ", " ", StormData$EVTYPE, useBytes = TRUE)
        if (length(grep("  ", StormData$EVTYPE, useBytes = TRUE)) == 0) {break}
}

StormData$STATE <- toupper(trimws(StormData$STATE))


# weather events classes  https://www.weather.gov/iwx/events

# Add a NEW column EVCLASS to StormData data set
StormData <- StormData %>% mutate(EVCLASS = character(1))

# weather event classification coding
# A - TORNADOs of all kinds 
# B - EXTREME TEMPERATURES on both end of the scale
# C - SNOWY & ICY conditions
# D - HIGH WINDS & STORM
# E - RAIN & FLOOD
# F - HAILing condition
# G - CLOUD condition that may lead to storm or tornado event
# H - FIRE 
# I - LIGHTNING
# X - EXCLUDED list - weather events to be excluded from processing
# Z - OTHER events not classified 

# Index 24, 25 & 26 of the ev.class array are the special designated classes.
# DO NOT ALTER the hard-designated values in index 24, 25 & 26, except the EXCLUSION keyword for weather events at index [24,2].
# Index 24 - EXCLUSION class, use for assigning weather events that are to be excluded from analytic processing. Change the search keyword of the weather event in the character string at index[24,2] as require.
# Index 25 - reserved.
# Index 26 - OTHER class, use for classifying weather events not classified by the regular class codes from index 1 to 23, and 25. Do NOT alter.
# DO NOT assign empty string value to the array cells, otherwise will result in unexpected outcome.
# DO NOT ALTER the sequence order of the default class codes and search keywords (array index 1 ~ 23).
ev.class <- array(dim = c(26,3))
ev.class[1,1] <- "A"        # CLASS CODE
ev.class[1,2] <- "TORNADO"  # CLASS LABEL
ev.class[1,3] <- "TORNADO"  # SEARCH KEYWORD
ev.class[2,1] <- "B"
ev.class[2,2] <- "EXTREME.TEMP"
ev.class[2,3] <- "HEAT|HOT|DRY|DROUGHT|COLD|CHILL|FROST|TEMP"
ev.class[3,1] <- "C"
ev.class[3,2] <- "SNOW.ICE"
ev.class[3,3] <- "SNOW|ICE|WINTER|BLIZZARD|FREEZ(E|ING)|SLEET|ICY"
ev.class[4,1] <- "D"
ev.class[4,2] <- "WIND.STORM"
ev.class[4,3] <- "WIND|STORM|GUSTNA|HURRICAN|TYPHOON|TURBULENCE|BURST|DUST|TSTM"
ev.class[5,1] <- "E"
ev.class[5,2] <- "RAIN.FLOOD"
ev.class[5,3] <- "FLOOD|WET|RAIN|SHOWER|HEAVY|PRECIPITAT"
ev.class[6,1] <- "F"
ev.class[6,2] <- "HAIL"
ev.class[6,3] <- "HAIL"
ev.class[7,1] <- "G"
ev.class[7,2] <- "CLOUD"
ev.class[7,3] <- "CLOUD|FUNNEL|WALL"
ev.class[8,1] <- "H"
ev.class[8,2] <- "WILD.FIRE"
ev.class[8,3] <- "FIRE"
ev.class[11,1] <- "I"
ev.class[11,2] <- "LIGHTNING"
ev.class[11,3] <- "LIGHTNING"
ev.class[24,1] <- "X"         # EXCLUSION class. Hard-designated. DO NOT ALTER.
ev.class[24,2] <- "EXCLUDED"  # class label. Hard-designated. DO NOT ALTER.
ev.class[24,3] <- "SUMMARY"   # EXCLUSION keyword, change as necessary.
ev.class[26,1] <- "Z"         # Hard-designated. DO NOT ALTER.
ev.class[26,2] <- "OTHER"     # class label. Hard-designated. DO NOT ALTER.
ev.class[26,3] <- "OTHER"     # OTHER class. DO NOT ALTER.


# 0. EXCLUDED - weather events to exclude
evclass.rowid <- grep(ev.class[24,3], StormData$EVTYPE)
evclass.evtyp <- StormData[evclass.rowid,]$EVTYPE
evclass <- cbind(evclass.rowid, evclass.evtyp)
StormData <- StormData %>% 
        mutate(EVCLASS = replace(EVCLASS, evclass.rowid, ev.class[24,1]))

for (i in 1:23) {
        if (!is.na(ev.class[i,1]) & !is.na(ev.class[i,3])) {
                evclass.rowid <- grep(ev.class[i,3], StormData$EVTYPE)
                evclass.rowid <- evclass.rowid[!(evclass.rowid %in% as.integer(evclass[,1]))]
                evclass.evtyp <- StormData[evclass.rowid,]$EVTYPE
                evclass.tmp <- cbind(evclass.rowid, evclass.evtyp)
                evclass <- rbind(evclass, evclass.tmp)
                StormData <- StormData %>%
                  mutate(EVCLASS = replace(EVCLASS, evclass.rowid, ev.class[i,1]))
        }
}

# 00. OTHER - weather events not classified by the regular class codes
evclass.rowid <- as.integer(row.names(StormData))
evclass.rowid <- evclass.rowid[!(evclass.rowid %in% as.integer(evclass[,1]))]
evclass.evtyp <- StormData[evclass.rowid,]$EVTYPE
evclass.tmp <- cbind(evclass.rowid, evclass.evtyp)
evclass <- rbind(evclass, evclass.tmp)
StormData <- StormData %>% 
        mutate(EVCLASS = replace(EVCLASS, evclass.rowid, ev.class[26,1]))

rm(evclass.tmp, evclass.evtyp, evclass.rowid, evclass)
```


```{r data.analytic}
par.def <- par(no.readonly = TRUE) # snapshot of default graphic parameters.

# all states of the USA (exclude off-shore affiliated territories).
states <- c(unique(state.fips$abb), "AK", "HI") 

# set up colour palette.
set.seed(2)
i <- sample.int(502, 256)
i <- sample(i, 66)
pal <- colors(TRUE)[i]

# assign ev.class.label to a character vector for simpler handling
ev.clvl <- ev.class[order(ev.class[1:23,1]),1:2]
N <- NROW(ev.clvl) - sum(is.na(ev.clvl[,1]))
pal <- pal[seq(length(pal), by = -2, length.out = N)]


# subset of StormData across the USA (exclude off-shore territories).
stormsub <- StormData %>% 
        select(EV_TIMESTAMP,EVCLASS,EVTYPE,LONGITUDE,LATITUDE,STATE,COUNTY,
               FATALITIES,INJURIES,PROPDMG,PROPDMGEXP,CROPDMG,CROPDMGEXP) %>%
        filter(STATE %in% states & !(EVCLASS %in% c("X","Z"))) %>% 
        mutate(STATE = factor(STATE),
               EVCLASS = factor(EVCLASS, labels = ev.clvl[1:N,2])) %>%
        arrange(EV_TIMESTAMP)
                # (!is.na(LONGITUDE) & !is.na(LATITUDE)) &
                #        (LONGITUDE != 0 & LATITUDE != 0)) %>%


# set up plotting layout and global graphics parameters 
layout(matrix(c(1,2,
                3,3,
                4,4), nrow = 3, byrow = TRUE))
par(las = 1, cex.axis = 0.65, ann = FALSE)

# aggregate weather events across the USA.
evcnt <- stormsub %>%
        select(STATE,EVCLASS,EVTYPE,FATALITIES,INJURIES,PROPDMG,CROPDMG) %>%
        group_by(EVCLASS) %>%
        count(EVCLASS) 

# plotting barchart for weather events aggregate across the USA.
par(mai = c(0.50,1.15,0.50,0.15))
barplot(evcnt$n, horiz = TRUE, border = NA, col = pal, names.arg = unique(evcnt$EVCLASS))

maxtik <- max(axTicks(1))
m <- barplot(evcnt$n, horiz = TRUE, plot = FALSE)
for (i in 1:length(m)) {
        v <- ifelse(evcnt$n[i] >= maxtik, -2.25, 0.25)
        text(evcnt$n[i], m[i], labels = formatC(evcnt$n[i], big.mark = ","), pos = 4,  offset = v, cex = 0.70)
}
mtext("Aggregate Frequency of Weather Events in USA (1950 ~ 2011)", font = 2, side = 3, line = -1.25, adj = 1, cex = 0.70)
mtext("(US Affiliated Territories Not Included)", side = 3, line = -2.00, adj = 1, cex = 0.65)

# aggregate weather events manifested in casualty and economic loss across the USA.
evcnt <- stormsub %>%
        select(STATE,EVCLASS,EVTYPE,FATALITIES,INJURIES,PROPDMG,CROPDMG) %>%
        filter(FATALITIES > 0 | INJURIES > 0 | PROPDMG > 0 | CROPDMG > 0) %>%
        group_by(EVCLASS) %>%
        count(EVCLASS) 

# plotting barchart for weather events aggregated across the USA that manifested in fatalities/injuries and properties/crops damages across the USA.
par(mai = c(0.50,0.15,0.50,1.15), yaxt = "n")
barplot(evcnt$n, horiz = TRUE, border = NA, col = pal, names.arg = unique(evcnt$EVCLASS))

maxtik <- max(axTicks(1))
m <- barplot(evcnt$n, horiz = TRUE, plot = FALSE)
for (i in 1:length(m)) {
        ifelse(evcnt$n[i] >= maxtik, v <- -2.25, v <- 0.25)
        text(evcnt$n[i], m[i], labels = formatC(evcnt$n[i], big.mark = ","), pos = 4,  offset = v, cex = 0.70)
}
mtext("Aggregate Frequency of Weather Events in USA (1950 ~ 2011)", font = 2, side = 3, line = -1.25, adj = 1, cex = 0.70)
mtext("Manifested in Casualty & Economic Loss", font = 2, side = 3, line = -2.00, adj = 1, cex = 0.70)
mtext("(US Affiliated Territories Not Included)", side = 3, line = -2.75, adj = 1, cex = 0.65)

par(cex.axis = 0.70, cex.lab = 0.80, ann = FALSE, yaxt = "s")

# weather events aggregate of individual state.
evcnt <- stormsub %>%
        select(STATE,EVCLASS,EVTYPE,FATALITIES,INJURIES,PROPDMG,CROPDMG) %>%
        group_by(EVCLASS, STATE) %>%
        count(EVCLASS, STATE) 

# plotting barchart for the occurrence of weather-events of each state.
par(mai = c(0.25, 1.15, 0.25, 1.15))
barplot(n ~ EVCLASS + as.character(STATE), data = evcnt, 
        horiz = FALSE, 
        border = NA,
        col = pal)
title(ylab = "Aggregate Frequency by Events")
mtext("Weather Events of Individual-State", font = 2, side = 3, line = -1.25, adj = 0, padj = 0, cex = 0.70)
mtext("(1950 ~ 2011)", side = 3, line = -2.00, adj = 0, padj = 0, cex = 0.65)

# aggregate weather events of individual state manifested in casualty and economic loss.
evcnt <- stormsub %>%
        select(STATE,EVCLASS,EVTYPE,FATALITIES,INJURIES,PROPDMG,CROPDMG) %>%
        filter(FATALITIES > 0 | INJURIES > 0 | PROPDMG > 0 | CROPDMG > 0) %>%
        group_by(EVCLASS, STATE) %>%
        count(EVCLASS, STATE) 

# plotting barchart for weather-events that triggered the occurrences of fatalities/injuries and properties/crops damages for each state.
par(mai = c(0.25, 1.15, 0.25, 1.15))
barplot(n ~ EVCLASS + as.character(STATE), data = evcnt, 
        horiz = FALSE, 
        border = NA,
        col = pal)
title(ylab = "Aggregate Frequency by Events")
mtext("Weather Events of Individual-State", font = 2, side = 3, line = -1.25, adj = 0, padj = 0, cex = 0.70)
mtext("Manifested in Casualty & Economic Loss", font = 2, side = 3, line = -2.00, adj = 0, padj = 0, cex = 0.70)
mtext("(1950 ~ 2011)", side = 3, line = -2.75, adj = 0, padj = 0, cex = 0.65)

par(par.def)

rm(i, v, maxtik, m, par.def, evcnt)
```

```{r house.keeping}
# detach("package:xts", unload = TRUE)
detach("package:dplyr", unload = TRUE)
detach("package:data.table", unload = TRUE)
```
